### 2016-09-02 ms/ms/ms.tex
added abstract to ms

\def\title{Increasing the tractability of occupancy models}
\def\author{Lauren C.\ Ponisio$^{1,2}$, Perry de Valpine$^1$}

@@

\begin{abstract}  
place holder
\end{abstract}

@@

Occupancy models are part of a larger class of models known as Hidden
Markov Models. For discrete Hidden Markov Models like occupancy models
where a species is either present or absent from a site, likelihood
calculation involves summing over the distribution of latent
states. Because estimating the effect of explanatory variables on site
occupancy or shared variation of in occupancy across species is often
of greatest interest to ecologsits
\citep[e.g.,][]{iknayan2014detecting}, the Hidden Markov Models are
embedded in a larger hierarchical model. In such cases, practitioners
may rely on Markov chain Monte Carlo (MCMC) to perform a Bayesian
analysis. Such models are computationally intensive, and large models
requiring hundreds or thousands of dimensions which require MCMC can
be computationally impractical. To ensure occupancy models are
accessible to practitioners, more efficient methods for estimating these
models are necessary.

%% model selection

>>>>

\def\title{Increasing the tractability of occupancy models}
\def\author{Lauren C.\ Ponisio$^{1,2}$, Nicholas Michaud$^1$, Perry de
  Valpine$^1$}

@@

\begin{abstract}  
  \begin{enumerate}
  \item occupancy models are everywhere, but model fitting and
    assessment are extrememly computationally intensive
  \item Using the NIMBLE package for R, we develop combined
    computational approaches to improve computational efficiency
    including user defined and automatic blocking of parameters for
    MCMC, filtering over latent states, and customized MCMC samplers
    for specific parameters. We test these approaches for three
    representative occupancy models of varying levels of complexity
    including a single species dynamic (i.e., multi-season) model with
    spatial auto correlation, and a mutli-species model. We also use
    calibrated posterior $p$-values to assess model fit.
  \item These computation approaches lead to an improvement in MCMC
    sampling efficiency, particularly with the more complex,
    mutli-species model.
  \item Ours results highlight the need for more customizable
    approaches to MCMC to fit and assess hierachical models in order
    to ensure occupancy models are accessible to practitioners. By
    implementing MCMC procedures and model assessment techniques open
    source software, we have made progress toward tis aim.
  \item \textit{Implications:}
  \end{enumerate}
\end{abstract}

@@

Occupancy models are part of a larger class of models known as Hidden
Markov Models. For discrete Hidden Markov Models like occupancy models
where a species is either present or absent from a site, likelihood
calculation involves summing over the distribution of latent
states. Because estimating the effect of explanatory variables on site
occupancy or shared variation of in occupancy across species is often
of greatest interest to ecologsits
\citep[e.g.,][]{iknayan2014detecting}, the Hidden Markov Models are
embedded in a larger hierarchical model. In such cases, practitioners
may rely on Markov chain Monte Carlo (MCMC) to perform a Bayesian
analysis. Such models are computationally intensive, and large models
requiring hundreds or thousands of dimensions which require MCMC can
be computationally intensive. In addition, fitting these models is
such a challenge that users often forgo adding any additional
computation to asses model fit. To ensure occupancy models are
accessible to practitioners, more efficient methods for fitting and
assessing these models are necessary.
